Volunteers in the Clouds: an Architecture for Low Cost and Potentially Massive Distributed Evolutionary Computation

### 2015-11-16 special-issue-ci/volunteer-clouds-ieee-ci-si.tex
Discussion of F15 experiment

\subsection{$F_{15}:\frac{D}{m}$-group Shifted and $m$-rotated Rastrigins Function}
The most common use of JavaScript is in the development of
asynchronous Web applications. For this purpose JavaScript has
been traditionally implemented as an interpreted language not designed for the
development of high performance systems. On the other hand, current JavaScript 
virtual machines (VMs) are closing the gap, for instance the Google V8 
is an engine specifically designed for fast execution of 
large JavaScript applications. In order to increase performance
V8 compiles JavaScript source code directly into machine code when it is first executed. 
There are no intermediate byte codes or an interpreter.
[\url{ https://developers.google.com/v8/design#mach_code}].
JavaScript has gained popularity in server-side development using runtime
environments such as Node.js and is also used in desktop and mobile applications.
To further evaluate how a current JavaScript implemantation could be used by a
computer scientist to develop complex optimization problems,
a benchmark function, which was provided by the CEC2010 Special Session on
Large-Scale Global Optimization [] is used in this section. 
The function is described next.

The basic Rastrigins function is separable[], and is defined as follows;
\begin{equation}
F_{rastrigin}(x)=\sum\limits_{i=1}^D [ x_{i}^{2}-10\cos(2\pi x_i)+10  ] 
\end{equation}
where $D$ is the dimension and $x = (x1, x2, · · · , x_{D})$ is a
$D$-dimensional row vector (i.e., a $1 × D$ matrix). Rastrigin’s function 
is a classical multimodal problem. It is difficult since the number of local
optima grows exponentially with the increase of dimensionality. To make it 
nonseparable, an orthogonal matrix is also used for coordinate rotation.
The rotated Rastrgin’s function is defined as follows:
\begin{equation}
F_{rot\_rastrigin}(x)=F_{rastrigin}(\textbf{z}), \textbf{z}= \textbf{x} * \textbf{M}
\end{equation}
Where $\textbf{M}$ is  a $D \times{D}$ orthogonal matrix. The benchmark
emulates real-world optimization problems that most likely will consist 
of different groups of parameters with strong dependencies within
but little interaction between the groups. This issue is reflected in the benchmark
problems by first randomly dividing the objective variables
into several groups, each of which contains a number of variables. The parameter $m$ 
is used to control the number of variables in each group and hence, defining the degree
of separability:
\begin{equation}
F_{15}(x)=\sum\limits_{k=1}^\frac{D}{m}F_{rot\_rastrigin}[\textbf{z}(P_{(k-1)*m+1}:P_{k*m})] 
\end{equation}
For the benchmark $D = 1000$, Group size $m = 50$, $\textbf{x} = (x_1, x_2, \ldots , x_{D})$ is the
candidate solution, $\textbf{o} = (o_1, o_2, \ldots  , o_{D})$ is the shifted global optimum,
$\textbf{z} = \textbf{x} - \textbf{o}$ is the shifted candidate solution and
finally $\textbf{P}$ is a random permutation of $\lbrace1, 2, \ldots  ,D\rbrace$.

This optimization problem was selected because is representative of the kind of
algorithms, data types and structures employed in large scale optimization problems. 
The function was implemented in both matlab and Java languages as part of  the
Test Suite of the companion competition. To give competitors an idea of the
computationa cost of the challenge, the runtime required for 10,000 function 
evaluations (FEs) was published for a particular configuration, where the Java
implementation took 7596ms and the Matlab version 1115ms. The whole experiment had
3,000,000 FEs as the termination condition. Using the Java implementation as
a guide, the function described above was implemented to compare the performence against
both implementations, again using the test of time required for 10,000 FEs. 
Before discussing the setup and performance results, certain details of
the implementation in JavaScript are presented, to give readers an idea of 
how practical and mature is the language for developing complex mathematical functions
as well as some limitations.

\begin{itemize}
\item {\em Randomize} Both Malab and Java implementations rely on a 
Java Randomizer library as a standardized implementation of the 
randomization of shift vectors and matrices used for the experiments.
The library included the generation of pseudorandom Gaussian of type
double and uniformly distributed integers. Generators used a single
{\tt long} seed. In order to implement the same functionality in
JavaScript, the library random-js was used, as there are inconsistencies 
in the implementation of {\tt Math.random()} between engines in the
number of bits used in the generation and results are non-deterministic. 
The library random-js, implements a Mersenne Twister algorithm producing 
consistent results across all JavaScript implementations assuming the same seed.
\item {\em Arrays} 
\item {\em Timing functions}
\item {\em Tools} 
\end{itemize}

The performance of the time required for 10,000 FEs is presented in Figure \ref{fig:f15_times}.
The test was run on an 3.7 GHz Quad-Core Intel Xeon E5 processor running OS X 10.10.5, 
using Java(TM) SE Runtime Environment (build 1.8.0\_25-b17), 
MatLab version R2015a for Mac OSX, Node FrameWork version 0.12.2, Google Chrome Version 46.0.2490.86 (64-bit) 
and Firefox version 41.0.2.

>>>>

\subsection{$F_{15}:\frac{D}{m}$-group Shifted and $m$-rotated Rastrigins Function}
The most common use of JavaScript is in the development of
asynchronous Web applications. For this purpose JavaScript has
been traditionally implemented as an interpreted language not designed for the
development of high performance systems. On the other hand, current JavaScript 
virtual machines (VMs) are closing the gap, for instance the Google V8 
is an engine specifically designed for the fast execution of 
large JavaScript applications. In order to increase performance
V8 compiles JavaScript source code directly into machine code when it is first executed. 
There are no intermediate byte codes or an interpreter.
[\url{ https://developers.google.com/v8/design#mach_code}].
JavaScript has gained popularity in server-side development using runtime
environments such as Node.js and is also used in desktop and mobile applications.
To further evaluate how a current JavaScript implemantation could be used by a
computer scientist to develop complex optimization problems,
a benchmark function, which was provided by the CEC2010 Special Session on
Large-Scale Global Optimization [] is used in this section. 
The function is described next.

The basic Rastrigins function is separable[], and is defined as follows;
\begin{equation}
F_{rastrigin}(x)=\sum\limits_{i=1}^D [ x_{i}^{2}-10\cos(2\pi x_i)+10  ] 
\end{equation}
where $D$ is the dimension and $x = (x1, x2, · · · , x_{D})$ is a
$D$-dimensional row vector (i.e., a $1 × D$ matrix). Rastrigin’s function 
is a classical multimodal problem. It is difficult since the number of local
optima grows exponentially with the increase of dimensionality. To make it 
nonseparable, an orthogonal matrix is also used for coordinate rotation.
The rotated Rastrgin’s function is defined as follows:
\begin{equation}
F_{rot\_rastrigin}(x)=F_{rastrigin}(\textbf{z}), \textbf{z}= \textbf{x} * \textbf{M}
\end{equation}
Where $\textbf{M}$ is  a $D \times{D}$ orthogonal matrix. The benchmark
emulates real-world optimization problems that most likely will consist 
of different groups of parameters with strong dependencies within
but little interaction between the groups. This issue is reflected in the benchmark 
by first randomly dividing the objective variables
into several groups, each of which contains a number of variables. The parameter $m$ 
is used to control the number of variables in each group and hence, defining the degree
of separability:
\begin{equation}
F_{15}(x)=\sum\limits_{k=1}^\frac{D}{m}F_{rot\_rastrigin}[\textbf{z}(P_{(k-1)*m+1}:P_{k*m})] 
\end{equation}
For the benchmark $D = 1000$, Group size $m = 50$, $\textbf{x} = (x_1, x_2, \ldots , x_{D})$ is the
candidate solution, $\textbf{o} = (o_1, o_2, \ldots  , o_{D})$ is the shifted global optimum,
$\textbf{z} = \textbf{x} - \textbf{o}$ is the shifted candidate solution and
finally $\textbf{P}$ is a random permutation of $\lbrace1, 2, \ldots  ,D\rbrace$.

This optimization problem was selected because is representative of the kind of
algorithm, data types and structures employed in large scale optimization problems. 
The function was implemented in both Matlab and Java languages as part of  the
Test Suite of the companion competition of the Special Session. To give competitors an idea of the
computational cost of the challenge, the runtime required for 10,000 function 
evaluations (FEs) for a particular configuration was published by organizers. The Java
implementation took 7596ms and the Matlab version 1115ms. The whole experiment had the value of
3,000,000 FEs as the termination condition. Using the Java implementation as
a guide, the function described above was implemented to compare the performence of JavaScript against
both implementations, again using as a test, the time required for doing a 10,000 FEs. 
Before discussing the setup and performance results, certain details of
the implementation in JavaScript are presented, to give readers an idea of 
how practical and mature is the language for developing complex mathematical functions
as well as some limitations.

\begin{itemize}
\item {\em Randomize} Both Malab and Java programs rely on a 
Java Randomizer library for the implementation of the 
randomization of shift vectors and matrices used for by the function.
The library included the generation of pseudorandom Gaussians of type
double and uniformly distributed integers. Generators used a single
{\tt long} seed. In order to implement the same functionality in
JavaScript, the library {\tt random-js} was used, as there are inconsistencies 
in the implementation of {\tt Math.random()} between engines and their results are non-deterministic. 
The library {\tt random-js}, implements a Mersenne Twister algorithm producing 
consistent results across all JavaScript VMs assuming the same seed.
\item {\em Timing functions} In order to compare the performance of 
each implementation, an accurate and consistent way of measuring 
the runtime is needed. In JavaScript the {\tt Date class} is often used 
to measure execution time, but has the limitaoun of having a maximum resolution given in miliseconds.
In order to measure the runtime intervals with a higher precission two functions
where used: for Node.js we used the native {\tt process.hrtime()} function
which returns a high-resolution measure in a {\tt [seconds, nanoseconds]} array.
For testing in the browser, the {\tt Performance.now()} function was used, this
function uses floating-point numbers with up to microsecond precision. 
A drawback found is that the {\tt Performance.now()} function is that is only implemented
in Firefox and Chrome browsers. Both functions are independent of the system clock. 
\item {\em Tools} An important consideration when chosing certain language to 
to write non-trivial programs is the availability of development tools. When developing
the experiments, we used a comprehensive set of developer tools available in both
browsers and desktop. These tools included package managers, debbugers, logging, 
network monitors, editors and even debugging of the multi-threaded execution of Web Workers.
A drawback found is that developer tools very from browser to browser, 
but they all have the same functionality.
\item {\em Data Types} JavaScript uses floating point numbers with a 
limited precision of 64 bits that partialy implement the functionality of
the {\tt StrictMath} library used by the Java implemantation. If more precision
is needed developers could use {\tt math.js}, an extensive math library with support
for matrices, big and complex numbers.
\end{itemize}

We found that the JavaScript ecosystem is mature enough for developing non-trivial 
code, but developers have to consider the differences betwwen VMs implementations
and also use certain libraries in order to increase the precision and repeatability of
algorithms. As for the performance of the JavaScript implementaion against Java and Matlab
we conducted first the basic test described earlier in a 3.7 GHz Quad-Core Intel Xeon E5 processor running OS X 10.10.5, 
using Java(TM) SE Runtime Environment (build 1.8.0\_25-b17), 
MatLab version R2015a for Mac OSX, Node FrameWork version 0.12.2, Google Chrome Version 46.0.2490.86 (64-bit) 
and Firefox version 41.0.2. The performance of the time required for 10,000 FEs is presented in Figure \ref{fig:f15_times}.
In our test the Matlab implementation had the best performance with an average of 935ms, followed by Java with
991ms, JavaScipt took 35\% more time than Java with 1,270ms using Node.js and 1279ms in Chrome. 
The results also show that there is not much difference between the Node.Js implemenation an that of a
browser. There is also not much difference between running the code in the main thread or as workers, but
when running the experiment in Firefox a dialog alerted of the script taking to long, showing that this
kind of processing is not intended for the main thread.
